---
title: Question 3 Code
author: Luke Dillon
output: html_document
---

```{r echo=FALSE}
library(tidyverse)
library(infer)
library(randomForest)
library(tidymodels)
library(modelr)
library(yardstick)
load("../../data/biomarker-clean.RData")
```

## Training/Testing Split
```{r splitData}
set.seed(101422)

biomarker_split <- biomarker_clean %>%
  initial_split(prop = 0.8)

training_biomarker <- training(biomarker_split)
testing_biomarker <- testing(biomarker_split)
```

Instead of splitting into training/testing at the end, we are doing it at the beginning. 
Everything else will be run as it was in the in-class analysis except we will be using the training split 
to run our t-tests and random forest.

```{r tTests}
test_fn <- function(.df) {
  t_test(.df,
    formula = level ~ group,
    order = c("ASD", "TD"),
    alternative = "two-sided",
    var.equal = FALSE
  )
}

ttests_out <- training_biomarker %>%
  # drop ADOS score
  select(-ados) %>%
  # arrange in long format
  pivot_longer(-group,
    names_to = "protein",
    values_to = "level"
  ) %>%
  # nest by protein
  nest(data = c(level, group)) %>%
  # compute t tests
  mutate(ttest = map(data, test_fn)) %>%
  unnest(ttest) %>%
  # sort by p-value
  arrange(p_value) %>%
  # multiple testing correction
  mutate(
    m = n(),
    hm = log(m) + 1 / (2 * m) - digamma(1),
    rank = row_number(),
    p.adj = m * hm * p_value / rank
  )
```

```{r randForest}
# store predictors and response separately
predictors <- training_biomarker %>%
  select(-c(group, ados))

response <- training_biomarker %>%
  pull(group) %>%
  factor()

# fit RF
set.seed(101422)
rf_out <- randomForest(
  x = predictors,
  y = response,
  ntree = 1000,
  importance = T
)

# check errors
rf_out$confusion
```

## Selecting Variables

```{r fuzzySelection}
ttest_adjp <- ttests_out %>%
  select(protein, p.adj)

rforest_gini <- rf_out$importance %>%
  as_tibble() %>%
  mutate(protein = rownames(rf_out$importance)) %>%
  select(protein, MeanDecreaseGini)

protein_scores <- ttest_adjp %>%
  left_join(rforest_gini, by = "protein")

protein_scores <- protein_scores %>% mutate(
  norm_gini = (MeanDecreaseGini - min(MeanDecreaseGini, na.rm = TRUE)) /
    (max(MeanDecreaseGini, na.rm = TRUE) - min(MeanDecreaseGini, na.rm = TRUE)),
  norm_pval = 1 - ((p.adj - min(p.adj, na.rm = TRUE)) /
    (max(p.adj, na.rm = TRUE) - min(p.adj, na.rm = TRUE)))
)

protein_scores <- protein_scores %>%
  mutate(combined_score = (norm_gini + norm_pval) / 2)

proteins_sstar <- protein_scores %>%
  slice_max(combined_score, n = 15) %>%
  pull(protein)
```

I decided to try to merge the two variable importance results (p-adjusted, and gini index)
I did this by putting them both in a tibble with their corresponding proteins
and then min-max normalizing both of then so that the best p-value is 1 and the worst is 0
fot the gini index the highest gini score is 1 and the lowest is 0

Then I added these two normalized variables together and divided by two to get their combined score
From this combined score, I selected an the top 15 (arbitrary number) proteins to use as predictors.

Here are the selected variables:

`r proteins_sstar`

```{r logisticRegression}
training_biomarker <- training_biomarker %>%
  select(group, any_of(proteins_sstar)) %>%
  mutate(class = (group == "ASD")) %>%
  select(-group)

testing_biomarker <- testing_biomarker %>%
  select(group, any_of(proteins_sstar)) %>%
  mutate(class = (group == "ASD")) %>%
  select(-group)

# fit logistic regression model to training set
fit <- glm(class ~ .,
  data = training_biomarker,
  family = "binomial"
)

# evaluate errors on test set
class_metrics <- metric_set(
  sensitivity,
  specificity,
  accuracy,
  roc_auc
)

testing_biomarker %>%
  add_predictions(fit, type = "response") %>%
  mutate(est = as.factor(pred > 0.5), tr_c = as.factor(class)) %>%
  class_metrics(
    estimate = est,
    truth = tr_c, pred,
    event_level = "second"
  )
```
